Decompositions of relative Chevalley groups

### 2015-10-08 paper.tex
Finished rewriting introduction

\title {Decompositions of relative Chevalley groups}
\keywords {Chevalley groups, principal congruence subgroups, stability for $\K_1$, bounded generation, subsystem factorizations, $\SL_2$-factorizations. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

The condition $\asr(I)$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable rank} of D.~Estes and J.~Ohm which does not depend on the ambient ring $R$ (see section~\ref{sec:rel-asr}).

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of relative Chevalley groups in terms of Tits generators.
\begin{thm}\label{thm:width} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis one has
%TODO: Now check the coefficients of \rk(\Phi) in the expression below
\begin{alignat*}{2}
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell,\rC_\ell, \\
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell,\rD_\ell;
\end{alignat*}
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type in some global field $k$ having a real embedding. 
Assume that $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\end{enumerate}
\end{thm}
Notice that a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} implies finiteness of $W(\E(\Phi, R, I), \mathcal{Z}(\Phi^+))$
for $R=\mathbb{Z}$ and $\Phi=\rA_\ell$.

Another application of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{LiebNikShaSL2}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies a similar factorization for the elementary subgroup $\E(\Phi, R)$ 
with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

In the linear and symplectic case the above results can be extended to even larger class of rings (which e.\,g. includes all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system satisfying the following assumptions.
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \rk(\Phi) - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell,\rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\rk(\Phi).
\end{array}\]
Then the principal congruence subgroup $\G(\Phi,R,I)$ can be presented as a product of at most $N$ copies of its (regularly embedded) subgroups isomorphic to $\SL_2(R, I)$.
\end{thm}

Yet another application of parabolic factorizations concerns subsystem factorizations of Chevalley groups formulated in terms of submaximal subgroups of type $\rA_\ell$.
In case when $R=I$ is a field this question was studied by N.~Nikolov in \cite{NikProdDecomp}. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

It is also classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, from the classical Dennis---Vaserstein theorem (see~\cite{Va69} and \cite[Lemma~2.1]{ST76})
it follows that $\SL_{n+1}(R, I)$ is a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.
Using another Dennis---Vaserstein type factorization we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$ then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ conjugates of (regularly embedded) subgroups of type $\rA_{\ell-1}$.
\end{thm}

>>>>

\title {Decompositions of relative Chevalley groups}
\keywords {Chevalley groups, principal congruence subgroups, stability for $\K_1$, bounded generation, $\SL_2$-factorizations. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

The condition $\asr(I) \leq d$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable range} condition of D.~Estes and J.~Ohm which depends only on the ideal $I$ (see section~\ref{sec:rel-asr}).

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\E(\Phi, R)$ in terms of Tits generators and is, thus, a quantitative analogue of \cref{prop:Stepanov}
and a relative analogue of the results of~\cite{CK83, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}.
\begin{thm}\label{thm:width} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type in some global field $k$ having a real embedding. 
Assume that $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis one has
%TODO: Now check the coefficients of \rk(\Phi) in the expression below
\begin{alignat*}{2}
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell,\rC_\ell, \\
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell,\rD_\ell;
\end{alignat*}
\end{enumerate}
\end{thm}
The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{Tavgen93, TavgenThesis}),
while for the second and the third assertion we give a direct proof.
Compare \cref{thm:width} also with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.

Another application of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{LiebNikShaSL2}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.

In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, in the context of \cref{thm:srRI1} one can immediately obtain the decomposition of $\E(\Phi, R, I)$ with $3|\Phi^+|$ factors of type $\rA_1$.
Our next result further extends this to even larger class of rings (which e.\,g. includes ideals of all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system satisfying the following assumptions.
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \rk(\Phi) - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell,\rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\rk(\Phi).
\end{array}\]
Then the principal congruence subgroup $\G(\Phi,R,I)$ can be presented as a product of at most $N$ copies of its (regularly embedded) subgroups isomorphic to $\SL_2(R, I)$.
\end{thm}

Our yet another application of parabolic factorizations concerns product decompositions of Chevalley groups formulated in terms of subgroups of type $\rA_\ell$ of submaximal rank.
Recall that in \cite{NikProdDecomp} N.~Nikolov proves the following result. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

It is classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, one immediately obtains from \cref{thm:DennisVaserstein} that $\SL_{n+1}(R, I)$ factors into a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.

Applying \cref{thm:DennisVaserstein} and generalizing (in the special case $\Phi=\rD_\ell$) Nikolov's \cite[Proposition~1]{NikProdDecomp} to arbitrary rings we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$ then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ conjugates of (regularly embedded) subgroups of type $\rA_{\ell-1}$.
\end{thm}
